Optimal Parameter Selection in Markov State Models for Biomolecular Conformational Dynamics

### 2013-11-26 paper/paper2.tex
added fig
changed style in the math stuff

\usepackage{amsmath}    % need for subequations
\usepackage{amssymb}    % for symbols
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external
                        % documents and URLs
\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
\usepackage{soul}

@@

The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis. With $s(X)$ as the function mapping conformations into the indicator function basis set, $s : \mathbb{R}^{3N} \rightarrow \{1, 2, \ldots, K\}$, the likelihood can be written as

\begin{align}
P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^{T} p_{s(x_i)}(x_{i})
\label{eq:like}
\end{align}

With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state transition probabilities, $T(s_i \rightarrow s_j)$, and so-called emission distributions of each state, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $x_t$ is within a certain state, $s(x_t)$.

@@

This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parameterizes a different statistical model. We propose two possible distributions: first, one may model the conditional distribution $P(x_t | s(x_t))$ as a uniform distribution over the (hyper)volume of $s(x_t)$, as shown in eq. \ref{eq:like_vol}. This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation. A more tractable alternative is a Gaussian emission model where $P(x_t | s(x_t))$ is modeled as multivariate normal, where the pairwise overlap between the emission distributions is treated as a truncation error, as shown in eq.  \ref{eq:like_vol} and used in the popular X-means algorithm.\cite{Pelleg2000Xmeans} The truncation error steps from the fact that the emission probability distributions from every state assign nonzero probability to each conformation in a dataset, while the parameterization procedure for the Markov transition probabilities assumes each conformation to be uniquely assigned to a single state. Relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.

@@

\begin{equation}
\label{eq:like_vol}
P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \cdot \prod_{i=0}^T \frac{1}{V_{s(x_{i})}}
\end{equation}

\begin{equation}
\label{eq:like_mvn}
\begin{split}
&P[x_{0...T-1}] dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
    &\cdot \prod_{i=0}^T \left(2 \pi\right)^\frac{d}{2} |\Sigma_{s(x_i)}|^{-\frac{1}{2}}  \exp\left(-\frac{1}{2} (x_i - \mu_{s(x_i)})' \Sigma_{s(x_i)}^{-1} (x_i - \mu_{s(x_i)})\right)
\end{split}
\end{equation}

where $\mu_{s(x_i)}$ is the cluster center assigned to $x_i$ and $\Sigma_{s(x_i)}$ is the corresponding covariance matrix. For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which conditions eq. \ref{eq:like_mvn} reduces to
\begin{equation}
\label{eq:like_mvn2}
\begin{split}
P[x_{0...T-1}] & dx^T = \prod_{i=0}^{T-1} T(s(x_i) \rightarrow s(x_{i+1})) \\
	&\cdot \prod_{i=0}^T \left(2 \pi \hat{\sigma}^2\right)^{-\frac{d}{2}} \exp\left(-\frac{||x_i - \mu_{s(x_i)}||^2}{2 \hat{\sigma}^2}\right)
\end{split}
\end{equation} where
\begin{equation}
\label{eq:mle_sigma}
\hat{\sigma}^2 = \frac{1}{T - k} \sum_{i=0}^T || x_i - \mu_{s(x_i)} ||^2
\end{equation} where $k$ is the number of states.

@@

For cross validation, we find it essential to use a dense prior during transition matrix estimation. Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes with maximum likelihood methods. For simplicity, we use the Jeffreys prior on each row of the transition matrix -- a Dirichlet prior with $\alpha=\frac{1}{2}$ -- which can be interpreted as a psuedocount of one half of an observed transition between each pair of states.

@@

\begin{figure}[h]
\centering
\includegraphics[width=2in]{figs_final/mull_cross_vert.png}
\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states. There are many methods that can be used to test the transferability of an MSM. We've plotted the AIC, BIC, and cross-validation scores for the M\"uller potential simulations. These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily.}
\label{fig:mullerlike}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs/kmeans_vor_like.png}
\caption{MSMs were built using K-Means clustering on the M\"uller potential simulations. Since K-Means optimizes an objective function that is related to the emission probability, our likelihood picks only a few states as the most likely model. Interestingly, the likelihood decreases as you introduce additional states. This behavior is due to the state-to-state likelihood decreased rapidly when adding additional states, which may be a result of the large number of states in the most stable well.}
\label{fig:kmeans_mull}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs/ww_logLs.png}
\caption{For models built using tICA, we were able to compute volumes as the dimensionality was low (A). These likelihoods were qualitatively similar to the MVN likelihoods (B). Because the MVN likelihood is easy to compute, we can use it to calculate the likelihood of a model clustered using the RMSD distance metric (C). Our likelihood indicates we should construct a model with 5000 states in the RMSD metric, whereas previous studies used a model with 26,000 states \cite{Lane2011Markov}.}
\label{fig:ww}
\end{figure}

>>>>

\usepackage{amsmath}    % need for subequations
\usepackage{amssymb}    % for symbols
\usepackage{graphicx}   % need for figures
\usepackage{verbatim}   % useful for program listings
\usepackage{color}      % use if color is used in text
\usepackage{subfigure}  % use for side-by-side figures
\usepackage{hyperref}   % use for hypertext links, including those to external
                        % documents and URLs
\usepackage[capitalise]{cleveref}   % use for referencing figures/equations
\usepackage{soul}
\usepackage{bm}

@@

The appropriate likelihood is instead a path action in phase space, on which the discrete states are merely an indicator function basis. With $s(X)$ as the function mapping conformations into the indicator function basis set, and $\{\mathbf{x}_t\}_{t=1}^N$ an observed trajectory, the likelihood can be written as:

\begin{equation}
\begin{split}
P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
\cdot \prod_{t=1}^{N} p_{s(\mathbf{x}_t)}(\mathbf{x}_{t}) \cdot d\mathbf{x}^N
\label{eq:like}
\end{split}
\end{equation}

With a discrete, non-overlapping state space, the likelihood of a sampled trajectory can be decomposed into a product of terms of two types: the state to state transition probabilities, $T(s_i \rightarrow s_j)$, and so-called emission distributions of each state, the conditional probability of observing a conformation at a given location in phase space given that the conformation, $\mathbf{x}_t$ is within a certain state, $s(\mathbf{x}_t)$.

@@

This emission distribution can be potentially modeled in multiple ways, each of which, in combination with a transition matrix, parameterizes a different statistical model. We propose two possible distributions: first, one may model the conditional distribution $P(\mathbf{x}_t | s(\mathbf{x}_t))$ as a uniform distribution over the (hyper)volume of $s(\mathbf{x}_t)$, as shown in eq. \ref{eq:like_vol}. This model has the advantage of appearing agnostic with respect to intra-state dynamics but is intractable in high dimensions when the states are defined by general polytopes, such as those produced by a data-driven Voronoi tessellation. A more tractable alternative is a Gaussian emission model where $P(\mathbf{x}_t | s(\mathbf{x}_t))$ is modeled as multivariate normal, where the pairwise overlap between the emission distributions is treated as a truncation error, as shown in eq.  \ref{eq:like_vol} and used in the popular X-means algorithm.\cite{Pelleg2000Xmeans} The truncation error steps from the fact that the emission probability distributions from every state assign nonzero probability to each conformation in a dataset, while the parameterization procedure for the Markov transition probabilities assumes each conformation to be uniquely assigned to a single state. Relaxing this approximation is possible with the forward-backward algorithm,\cite{Rabiner1989Tutorial} but is beyond the scope of this work.

@@

\begin{equation}
\label{eq:like_vol}
\begin{split}
P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
	\cdot \prod_{t=1}^N \frac{1}{V_{s(\mathbf{x}_{t})}} \cdot d\mathbf{x}^N
\end{split}
\end{equation}

\begin{equation}
\label{eq:like_mvn}
\begin{split}
&P\left[\{\mathbf{x}_t\}_{t=1}^N\right] d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
	&\;\;\cdot \prod_{t=1}^N\Bigg[ \frac{1}{(2 \pi)^\frac{d}{2} |\Sigma_{s(\mathbf{x}_t)}|^\frac{1}{2}} \\
    &\;\;\cdot \exp\left(-\frac{1}{2} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})^T \Sigma_{s(\mathbf{x}_t)}^{-1} (\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)})\right)\Bigg] \cdot d\mathbf{x}^N
\end{split}
\end{equation}

where $\bm{\mu}_{s(\mathbf{x}_t)}$ is the cluster center assigned to $\mathbf{x}_t$ and $\Sigma_{s(\mathbf{x}_t)}$ is the corresponding covariance matrix. For simplicity, we follow \citet{Pelleg2000Xmeans} and employ a spherical Gaussian model with a single shared variance parameter across the states estimated by maximum likelihood, under which conditions eq. \ref{eq:like_mvn} reduces to
\begin{equation}
\label{eq:like_mvn2}
\begin{split}
P\left[\{\mathbf{x}_t\}_{t=1}^{N}\right] & d\mathbf{x}^N = \prod_{t=1}^{N-1} T(s(\mathbf{x}_t) \rightarrow s(\mathbf{x}_{t+1})) \\
	&\cdot \prod_{t=1}^N \left(2 \pi \hat{\sigma}^2\right)^{-\frac{d}{2}} \exp\left(-\frac{||\mathbf{x}_t - \bm{\mu}_{s(\mathbf{x}_t)}||^2}{2 \hat{\sigma}^2}\right)
\end{split}
\end{equation} where
\begin{equation}
\label{eq:mle_sigma}
\hat{\sigma}^2 = \frac{1}{T - k} \sum_{i=0}^T || x_i - \mu_{s(x_i)} ||^2
\end{equation} where $k$ is the number of states.

@@

For cross validation, we find it essential to use a dense prior during transition matrix estimation. Because the maximum likelihood transition matrix fit on a subset of the data assigns probability zero to unobserved transitions, we observe cross validation log likelihoods of negative infinity in all but the most data-rich regimes with maximum likelihood methods. We used a dirichlet prior on each entry of the transition matrix, which

For simplicity, we use the Jeffreys prior on each row of the transition matrix -- a Dirichlet prior with $\alpha=\frac{1}{2}$ -- which can be interpreted as a psuedocount of one half of an observed transition between each pair of states.

@@

\begin{figure}[h]
\centering
\includegraphics[width=3in]{figs_final/mull_likes.png}
\caption{In models built using between 50 and 1,000 states with the $k$-centers algorithm, the log likelihood function increased quickly and plateaued at approximately 300 states. There are many methods that can be used to test the transferability of an MSM. We've plotted the AIC, BIC, and cross-validation scores for the M\"uller potential simulations. These results indicate that, at least for this dataset, the BIC and AIC may penalize the number of states too heavily.}
\label{fig:mullerlike}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=2.5in]{figs_final/kmeans_like2_lbl.png}
\caption{MSMs were built using K-Means clustering on the M\"uller potential simulations. Since K-Means optimizes an objective function that is related to the emission probability, our likelihood picks only a few states as the most likely model. Interestingly, the likelihood decreases as you introduce additional states. This behavior is due to the state-to-state likelihood decreased rapidly when adding additional states, which may be a result of the large number of states in the most stable well.}
\label{fig:kmeans_mull}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/ww_likes.png}
\caption{For models built using tICA, we were able to compute volumes as the dimensionality was low (A). These likelihoods were qualitatively similar to the MVN likelihoods (B). Because the MVN likelihood is easy to compute, we can use it to calculate the likelihood of a model clustered using the RMSD distance metric (C). Our likelihood indicates we should construct a model with 5000 states in the RMSD metric, whereas previous studies used a model with 26,000 states \cite{Lane2011Markov}.}
\label{fig:ww}
\end{figure}
